Parsing Postfix log files

### 2007-12-06 doc/logparser.tex
Improved other parsers review; it's more compare and contrast than it was. Picked Postfix Log Parser as the name. Added parser background section, introducing the rules/algorithm split. Hyperref doesn't work after upgrading to unstable, so it's disabled for now.

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
\newcommand{\parsername}{ASO}
\newcommand{\parsernames}{ASO's}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib]{tocbibind}

@@

% When creating a \PDF{} make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}
\usepackage{lastpage}

@@

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible to achieve a high degree of accuracy in understanding
    the logs, and thus reconstructing the actions taken by Postfix to
    generate the logs.  This paper describes the creation of a parser,
    documenting the parsing algorithm and rules, explaining the
    difficulties encountered and the solutions developed, with reference to
    an implementation which stores data gleaned from the logs in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.

@@

XXX THINK OF A NAME\@.

@@

@@

@@

XXX WHERE SHOULD THE NEXT PARAGRAPH BE PLACED\@?

Separating the rules and algorithm makes it possible to parse new log lines
without modifying the core parsing algorithm.  Although this may seem like
a trivial point, is it substantially more difficult to understand a
program's entire parsing algorithm, identify the correct location to
change, and make the appropriate changes without adversely affecting
parsing, particularly as there may be edge cases which are not immediately
obvious.\footnote{See
section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
which occurs only four times in 93 log files tested.}  Requiring changes to
the parsing algorithm also complicates upgrades, as the changes must be
preserved during the upgrade, and may clash with changes made by the
developer.  \parsername{} allows the user to add new rules to the database
without changing the parsing algorithm, unless the new log lines to be
parsed require functionality not already provided by the algorithm.  If the
new log lines do require new functionality, new actions can be added to the
parser without modifying existing actions or other parts of the algorithm;
only in the rare case that the new actions require support from other
sections of the code will more extensive changes be required.

@@

XXX COMPARE AND CONTRAST RATHER THAN REVIEW\@.

@@

        AWStats will produce simple graphs for many different services, but
        supporting many different services without special purpose code
        restricts it to supporting the \LCD{}.  The data it will extract
        from an \MTA{} log file is limited in comparison to \parsername{}:
        \newline \tab{} time2, email, email\_r, host, host\_r, method, url,
        code and bytesd.\newline  There does not seem to be an explanation
        of any of those fields in the documentation (\parsername{} provides
        copious documentation).  AWStats has no special purpose code to
        deal with the intricacies of Postfix logs, in fact it operates by
        coercing Postfix logs into Apache\footnote{The Apache web server is
        the most popular HTTP server in use for the past 10 years; more
        information is available at \url{http://httpd.apache.org/}.} format
        log files, for analysis by AWStats' HTTP log file parser.  The
        converting parser only deals with a small portion of the log lines
        generated by Postfix, silently skipping those it cannot deal with,
        and does not distinguish between different types of rejection.
        Although though it does correlate log lines by queueid, it does not
        deal with any of the other complications described in this
        document.

@@

    \item [Log analyser --- throughput monitor] This utility tracks the
        number of events which occurred over a particular time and warns if
        the frequency of events passes a certain threshold.  It's designed
        to provide real time alerts when dictionary attacks, mail loops or
        similar problems occur. It doesn't attempt to extract or save data,
        correlate log lines, or any of the more advanced tasks described in
        this document, because it is not designed to do so.

@@

        Anteater doesn't have any English documentation so it's difficult,
        nigh impossible, to accurately comment on what analysis is
        performs.  It did not run successfully when tested, and its parsing
        would certainly be out of date as Postfix has evolved considerably
        since this tool was last updated (November 2003).  As it neither
        ran successfully nor has documentation the author can read a
        detailed review cannot be provided.

@@

        YAALA uses a plugin based system to analyse log files and produce
        \HTML{} output reports.  The core code is merely 153 lines long, as
        all the parsing and report generation is handled by modules.  YAALA
        supports storing previously gathered data; it dumps its internal
        data structures using Perl's Storable module~\cite{perl-storable},
        so with some knowledge of the data structure created by the module
        responsible for parsing Postfix log files it should be possible to
        extract data with an external program.  Using YAALA as a base would
        be only slightly less work than starting from scratch, as both
        input and output modules would need to be written specially; it may
        even be more work to implement the parser within the constraints of
        YAALA\@.  YAALA bundles a Postfix parser, which extracts the
        following fields from specific log lines:

@@

        YAALA provides an \SQL{} like mini-language used when generating
        the reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \tab{} \texttt{requests BY file WHERE host
        =\~{} Google}. \newline The mini-language is quite limited and
        cannot be used to extract data for external use, merely to create
        reports.

        YAALA as distributed does not attempt the in-depth parsing
        explained in this document, and implementing such parsing within
        the confines of YAALA would be at least as much work as
        implementing it separately.

        In summary YAALA provides a Postfix parser which only handles the
        most common Postfix log lines, provides reasonably flexible report
        generation from the limited data extracted, but no facilities to
        extract data for use in other tools.

@@

        Lire is a general purpose log parser supporting many different
        types of log file.  Its Postfix parser extracts the following data
        from Postfix logs: \textit{The email servers' reports will show you
        the number of deliveries and the volume of email delivered by day,
        the domains from which you receive or send the most emails, the
        relays most used, etc.\/}; notably rejections are not mentioned or
        handled.  It supports multiple output formats (text, \HTML{},
        \PDF{} and Excel 95).  Logs can be parsed, the extracted data
        saved, and reports generated from it later.  Accessing this data
        from another application is undocumented; given the source code it
        should be possible, with sufficient time and effort, to access the
        data.

        Like AWStats and Logrep, Lire attempts to correlate log lines by
        queueid, but not by \pid{}, so the complete list of recipients for
        a mail should be available; however its parser extracts only part
        of the available data and makes no attempt to deal with the other
        complications described in sections~\refwithpage{complications}
        and~\refwithpage{additional complications}.  It stores data, but
        not in an easily accessible format, and generates reports in
        multiple formats, but those reports do not appear to be
        customisable.

@@

        Logrep extracts the following fields from Postfix logs:

@@

        Fields are correlated based on the queueid (referred to as
        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
        is error prone: empty fields are saved when the log line doesn't
        match the \regex{}, though it appears that they will not overwrite
        existing data.  It also counts the number of log lines parsed and
        skipped.  Most notably rejections are completely ignored, making it
        unsuitable for the purposes of this project.  It doesn't attempt to
        address any of the complications in
        sections~\refwithpage{complications} and~\refwithpage{additional
        complications} except for correlating by queueid.

        Logrep does not come with any documentation, though some scant
        documentation is available on its website.  It requires a web
        browser to interact with it, ensuring that automated log processing
        will be difficult.  Sadly all attempts to use Logrep failed, as it
        was unable to access the log files selected; this appears to be a
        bug rather than operator error.\footnote{If it is caused by
        operator error, the interface needs improvement as the (minimal)
        instructions were followed as closely as possible.}  As parsing
        failed it wasn't possible to review the reports Logrep can generate
        (available in \HTML{} only), nor to examine the (undocumented)
        format in which it can save extracted data for subsequent reuse.

>>>>

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.
\usepackage{url}
\newcommand{\parsername}{\PLP{}}
\newcommand{\parsernames}{\PLP{}'s}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib]{tocbibind}

@@

% XXX THIS DOESN'T WORK AFTER PACKAGE UPGRADES :(
% When creating a \PDF{} make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
%\usepackage{hyperref}
%\hypersetup{
%    pdftitle    = {Parsing Postfix log files},
%    pdfauthor   = {John Tobin},
%    final       = true,
%    pdfborder   = {0 0 0},
%%    breaklinks  = true,
%}
%\usepackage{breakurl}
\usepackage{lastpage}

@@

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible to achieve a high degree of accuracy in understanding
    the logs, and thus reconstructing the actions taken by Postfix to
    generate the logs.  This paper describes the creation of a parser,
    documenting the parsing algorithm, explaining the difficulties
    encountered and the solutions developed, with reference to an
    implementation which stores data gleaned from the logs in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.

@@

XXX NOW SAVING THE SIZE OF MAILS\@.

@@

\label{postfix background}

@@

\subsection{Parser design}

\label{parser design}

It should be clear from the earlier Postfix background
(section~\refwithpage{postfix background}) that logs produced by Postfix
are not fixed; they vary widely from host to host, depending on the set of
restrictions chosen by the administrator.  With this in mind, one of the
parser's design aims was to make adding new rules as easy as possible, to
enable administrators to properly parse their log lines.  The main design
decision which enables this is the separation of rules from the parsing
algorithm, and secondly the division of the algorithm into discrete
actions.

Separating the rules from the algorithm makes it possible to parse new log
lines without modifying the core parsing algorithm.  Although this may seem
like a trivial point, is it substantially more difficult to understand a
program's entire parsing algorithm, identify the correct location to
change, and make the appropriate changes without adversely affecting
parsing, particularly as there may be edge cases which are not immediately
obvious.\footnote{See
section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
which occurs only four times in 93 log files tested.}  Requiring changes to
the parsing algorithm also complicates upgrades, as the changes must be
preserved during the upgrade, and may clash with changes made by the
developer.  \parsername{} allows the user to add new rules to the database
without changing the parsing algorithm, unless the new log lines to be
parsed require functionality not already provided by the algorithm.  If the
new log lines do require new functionality, new actions can be added to the
parser without modifying existing actions or other parts of the algorithm;
only in the rare case that the new actions require support from other
sections of the code will more extensive changes be required.

@@

@@

@@

        AWStats will produce simple graphs for many different services, but
        supporting many different services without special purpose code
        restricts it to supporting the \LCD{}.  The data it will extract
        from an \MTA{} log file is limited in comparison to \parsername{}:
        \newline \tab{} time2, email, email\_r, host, host\_r, method, url,
        code and bytesd.\newline  There does not seem to be an explanation
        of any of those fields in the documentation (\parsername{} provides
        copious documentation).  AWStats has no special purpose code to
        deal with the intricacies of Postfix logs, in fact it operates by
        coercing Postfix logs into Apache\footnote{The Apache web server is
        the most popular HTTP server in use for the past 10 years; more
        information is available at \url{http://httpd.apache.org/}.} format
        log files, for analysis by AWStats' HTTP log file parser.  The
        converting parser only deals with a small portion of the log lines
        generated by Postfix, silently skipping those it cannot deal with,
        and does not distinguish between different types of rejection;
        there is no easy way to extend it to handle new log lines.
        Although though it does correlate log lines by queueid, it does not
        deal with any of the other complications described in this
        document.

@@

    \item [Log analyser --- throughput monitor] This utility tracks the
        number of events which occurred over a particular time and warns if
        the frequency of events passes a certain threshold.  It's designed
        to provide real time alerts when dictionary attacks, mail loops or
        similar problems occur. It doesn't attempt to extract or save data,
        correlate log lines, or any of the more advanced tasks described in
        this document, because it is not designed to do so.  The user must
        construct the \regexs{} to match significant input lines before it
        will parse any logs.  This program was not reviewed in further
        detail because its aims are so far removed from the aims of this
        project.

@@

        Anteater doesn't have any English documentation so it's difficult,
        nigh impossible, for this author to accurately comment on what
        analysis it performs.  It did not run successfully when tested, and
        its parsing would certainly be out of date as Postfix has evolved
        considerably since this tool was last updated (November 2003).  As
        it neither ran successfully nor has documentation the author can
        read a detailed review cannot be provided.

@@

        YAALA uses a plugin based system to analyse log files and produce
        \HTML{} output reports, with all the parsing and report generation
        handled by modules.  Using YAALA as a base would be only slightly
        less work than starting from scratch, as both input and output
        modules would need to be written specially; it may even be more
        work to implement the parser within the constraints of YAALA\@.
        YAALA supports storing previously gathered data using Perl's
        Storable module~\cite{perl-storable}, so with enough knowledge of
        the data structure YAALA stores it should be possible to extract
        data with another Perl program also using the Storable module;
        \parsername{} uses a well documented database which is accessible
        from the majority of programming languages.

        YAALA provides a Postfix parser which extracts the following fields
        from specific log lines:

@@

        YAALA extracts most of the fields \parsername{} does, but it does
        not maintain separate counters for each restriction like
        \parsername{}, so its data cannot be used for optimisation, testing
        or understanding of restrictions.  YAALA's Postfix parser does not
        deal with the complications explained in this document, though it
        does correlate log lines by queueid.

        YAALA provides a mini-language based on \SQL{} that is used when
        generating reports; sample reports can be seen
        at~\cite{yaala-samples}.  Example query: \newline \tab{}
        \texttt{requests BY file WHERE host =\~{} Google} \newline The
        mini-language is quite limited and cannot be used to extract data
        for external use, merely to create reports.

        In summary YAALA provides a Postfix parser which only handles the
        most common Postfix log lines, provides reasonably flexible report
        generation from the limited data extracted, but has no facilities
        to extract data for use in other tools.

@@

        Lire is a general purpose log parser supporting many different
        types of log file.  Its Postfix parser extracts the following data
        from Postfix logs: \textit{The email servers' reports will show you
        the number of deliveries and the volume of email delivered by day,
        the domains from which you receive or send the most emails, the
        relays most used, etc.\/}; notably rejections are not mentioned or
        dealt with, and unlike \parsername{} there is no facility to extend
        the parser.  It supports multiple output formats for generated
        reports (text, \HTML{}, \PDF{} and Excel 95) but the reports do not
        appear to be customisable; \parsername{} does not produce any
        reports.  Lire supports saving extracted data for later report
        generation, but accessing this data from another application is
        undocumented; given the source code it should be possible, with
        sufficient time and effort, to access the data from an external
        program.

        Like AWStats and Logrep, Lire attempts to correlate log lines by
        queueid, but not by \pid{}, so the complete list of recipients for
        a mail should be available; however its parser extracts only part
        of the available data and makes no attempt to deal with the other
        complications described in sections~\refwithpage{complications}
        and~\refwithpage{additional complications}.  \parsername{} uses an
        \SQL{} database to make accessing the extracted data as easy as
        possible.

@@

        Logrep extracts roughly half the fields \parsername{} does:

@@

        It also counts the number of log lines parsed and skipped.  Log
        lines are correlated based on the queueid (referred to as
        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
        is error prone: empty fields are saved when the log line doesn't
        match the \regex{}, though it appears that they will not overwrite
        existing data.  Most notably rejections are completely ignored,
        making it unsuitable for the purposes of this project.  It doesn't
        attempt to address any of the complications in
        sections~\refwithpage{complications} and~\refwithpage{additional
        complications} except for correlating by queueid.

        Logrep does not come with any documentation, though some scant
        documentation is available on its website (\parsername{} provides
        copious documentation).  It requires a web browser to interact with
        it, ensuring that automated log processing will be difficult,
        whereas automated processing is a key part of \parsernames{}
        design.  Sadly all the author's attempts to use Logrep failed, as
        it was unable to access the log files selected; this appears to be
        a bug rather than operator error.\footnote{If it is caused by
        operator error, the interface needs improvement as the (minimal)
        instructions were followed as closely as possible, and multiple
        attempts were made.}  As parsing failed it wasn't possible to
        review the reports Logrep can generate (available in \HTML{} only),
        nor to examine the (undocumented) format in which it can save
        extracted data for subsequent reuse.
